
# Visaulizing Florida Forever Project Accumulations through Manually parsed data

Ryan Anderson
Feb 13, 2026

This notebook attempts to create a proof of concept for how we might try to see
how a project has evolved over time, in this case we can set the boundary to a Florida Forever
BOT project, each of which has a linked PDF containing descriptions of how the project
evolved over time.

As a proof of concept, I parsed out bombing range ridge data and turned it into a usable table.

Here's the sort of visualization we might make.

```{r setup}
# clear env
rm(list=ls())

# load libs
library(tidyverse)
library(here)
library(janitor)

# load data
# data_raw/ff_bot_project_timelines_manually_parsed.csv
ff_bot_timelines <- read_csv(here("data_raw", "ff_bot_project_timelines_manually_parsed.csv")) %>%
  clean_names()

bombing_range_ridge <- ff_bot_timelines %>%
  filter(project_name == "Bombing Range Ridge")

```


Note: this data file will continue to be updated. The current version can be downloaded from the project GDrive: https://docs.google.com/spreadsheets/d/1eNoudZBucQE_7YirpqEpINCTQ28U_FTcMPwfAjc_AX0/edit?usp=sharing

Data fields
[1] "project_name"                            "fdep_report_link"                        "year"                                   
[4] "acres_acquired"                          "acquisition_type"                        "acres_of_project_boundary_expanded"     
[7] "intended_acquisition_type_for_expansion" "notes"              

Let's compute cumulative fields.
Let's get cumulative acres acquired and cumulative project boundary acres.

```{r}
bombing_range_ridge <- bombing_range_ridge %>%
  arrange(year) %>%
  mutate(cumulative_acres_acquired = cumsum(acres_acquired),
         cumulative_project_boundary_expanded = cumsum(acres_of_project_boundary_expanded))
```

Now let's do a quick pivot and make a double line graph (one for project boundary cumulative and the other for project acres acquired cumulative)

```{r}

# Pivot data to long format for plotting
bombing_range_ridge %>%
  select(year, cumulative_acres_acquired, cumulative_project_boundary_expanded) %>%
  pivot_longer(cols = c(cumulative_acres_acquired, cumulative_project_boundary_expanded),
               names_to = "metric",
               values_to = "cumulative_acres") %>% 
  ggplot(aes(x = year, y = cumulative_acres, color = metric)) +
  geom_line(size = 1) +
  labs(
    title = "Bombing Range Ridge",
    x = "Year",
    y = "Cumulative Acres",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


