
# Mass generation of curve plots

Ryan Anderson
March 12, 2026

In this notebook, I provide code to generate a BUNCH of plots. Like a BUUUUNCH.

The raw data can be downloaded from our drive at this link
https://drive.google.com/drive/folders/1VZlUSSH7tjnSOgBInS4itMEWe7_SHbU-?usp=drive_link

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r}
# clear env
rm(list=ls())

# Global parameter variables
GENERATE_CUMULATIVE_PLOTS <- FALSE
GENERATE_NON_CUMULATIVE_YEARLY_PLOTS <- TRUE

# load libraries
library(tidyverse)
library(here)
library(janitor)
library(readxl)
```


```{r}
# Load Data
# get the joined dataset
#data_raw/FFBOT_Spatial_Join_Acq-Proj.xlsx
ff_joined <- read_excel(here("data_raw", "FFBOT_Spatial_Join_Acq-Proj.xls")) %>%
  clean_names()

# Let's load up the ffbot dataset
ff_projects <- read_csv(here("data_raw", "FNAI_Spatial_Data_Attr_Tables", "ffbot_2025_12.csv")) %>%
  clean_names()
```

**Column Names**

 [1] "objectid"       "join_count"     "target_fid"     "id"             "par_name"       "prj_name"       "county"         "acq_type"       "close_date"    
[10] "ff_pur_pr1"     "ff_pur_pr2"     "oth_pur_pr"     "ff_funds_s"     "prc_ff"         "ff_ac_pur"      "acres"          "ma_name"        "ma_id"         
[19] "macode_bcd"     "man_agncy"      "in_flma"        "comments"       "dig_com"        "citation"       "in_poly_fid"    "sim_pgn_flag"   "max_simp_tol"  
[28] "min_simp_tol"   "site_id"        "sitename"       "short_name"     "short_site"     "project_cd"     "project_name"   "ff_cat"         "ff_rank"       
[37] "prisec_acres"   "gis_acres"      "sitedesc"       "boundjust"      "protcom"        "macom"          "site_relations" "sitecom"        "fdep_url"      
[46] "shape_length"   "shape_area"    



```{r}
# First, let's just get the sitename and shape_area from the ff_projects dataset
ff_projects_sitnames_areas <- ff_projects %>%
  select(sitename, shape_area) %>%
  rename(project_boundary_acres = shape_area)

# Next, let's select a subset of the ff_joined that includes the date, area (via shape_area), and the sitename (which is the primary key we want to use to link the datasets)
ff_joined_subset <- ff_joined %>%
  select(close_date, shape_area, sitename) %>%
  rename(acquisition_acres = shape_area)

# Now let's perform a left join to bring in the project boundary acres from the ff_projects dataset
ff_joined_with_boundaries <- ff_joined_subset %>%
  left_join(ff_projects_sitnames_areas, by = "sitename")

# Let's get a computed field that indicates the proportion of land within the project boundaries that the acquisition covers.
ff_joined_with_boundaries <- ff_joined_with_boundaries %>%
  mutate(proportion_of_project_acquired = acquisition_acres / project_boundary_acres)
```

Next let's get a field that cumulatively shows the total proportion of project acquired over time. This requires the processing of a time variable. So let's first convert the close_date field to a date format and then arrange the data by project and date.

```{r}
ff_joined_with_boundaries <- ff_joined_with_boundaries %>%
  mutate(close_date = as.Date(close_date, format = "%Y-%m-%d")) %>%
  arrange(sitename, close_date) %>%
  group_by(sitename) %>%
  mutate(cumulative_proportion_acquired = cumsum(proportion_of_project_acquired))
```


## Generating accumulation plots

So now I want to do something a little wild. We want to be able to see all our accumulation plots together so we can start to suss out some patterns.

So I'm going to try generating a bunch of them in a loop, and then outputting them to a folder.

First, let's make a helper function to create the accumulation plot. It should take in a sitename

```{r}
create_accumulation_plot <- function(sitename_param) {
  project_data <- ff_joined_with_boundaries %>%
    filter(sitename == sitename_param)
  
  p <- ggplot(project_data, aes(x = close_date, y = cumulative_proportion_acquired)) +
    geom_line() +
    geom_point() +
    labs(title = paste(sitename_param),
         x = "Close Date",
         y = "Cumulative Proportion Acquired") +
    theme_minimal()
  
  return(p)
}

# Test it out
test_plot <- create_accumulation_plot("Pal-Mar Florida Forever BOT Project")
test_plot
```
Now let's get an array of all sitenames and loop through it, generate the plot, then save to a folder called 
data_outputs/ff_accumulation_plots

But with one catch: don't do any site that has only a single data point or wo

## Creating all cumulative plot files

```{r}
# Get all unique sitenames of sites that exist in ff_joined_with_boundaries
# ie all sites that have had at least one piece of an acquisition occur within its bounds.
all_sitenames <- unique(ff_joined_with_boundaries$sitename)
```

```{r}
### Creating cumulative plots
# Loop through sitenames, create plot, and save
if(GENERATE_CUMULATIVE_PLOTS) {
  for (sitename_i in all_sitenames) {
    # if the number of data points for this sitename is less than or equal to 1, skip it
    if (nrow(ff_joined_with_boundaries %>% filter(sitename == sitename_i)) <= 2) {
      next
    }
  
    plot <- create_accumulation_plot(sitename_i)
    filename <- paste0(sitename_i, ".png")
    
    # change slashes to dashes to avoid creating new folders
    filename <- str_replace_all(filename, "/", "-")
    filepath <- here("data_outputs", "ff_accumulation_plots", filename)
    ggsave(filename = filepath, plot = plot, width = 8, height = 6)
  }
}

```


## Creating Non-Cumulative Yearly Plots -- Helper function

```{r}
# let's make a similar helper function that will take in a sitename and produce
# a non-cumulative plot of the proportion of project acquired year-by-year
create_yearly_noncum_plot <- function(sitename_param) {
  site_data <- ff_joined_with_boundaries %>%
    filter(sitename == sitename_param)
  
  # obtain year-by-year values by grouping by year and summing the proportion acquired for each year
  site_data_yearly <- site_data %>%
    group_by(year = lubridate::year(close_date)) %>%
    summarize(yearly_proportion_acquired = sum(proportion_of_project_acquired, na.rm = TRUE))
  
  # now let's make the plot
  p <- ggplot(site_data_yearly, aes(x = year, y = yearly_proportion_acquired)) +
    geom_line() +
    geom_point() +
    labs(title = paste(sitename_param),
         x = "Year",
         y = "Yearly Proportion Acquired") +
    theme_minimal()
  
  # redo as a bar plot
  # p <- ggplot(site_data_yearly, aes(x = year, y = yearly_proportion_acquired)) +
  #   geom_bar(stat = "identity") +
  #   labs(title = paste(sitename_param),
  #        x = "Year",
  #        y = "Yearly Proportion Acquired") +
  #   theme_minimal()
  
  
  
  return(p)
 
}

# Test
test_yearly_noncum_plot <- create_yearly_noncum_plot("Pal-Mar Florida Forever BOT Project")
test_yearly_noncum_plot

```

## Creating all non-cumulative yearly plot files

```{r}
# Loop through sitenames, create plot, and save
if(GENERATE_NON_CUMULATIVE_YEARLY_PLOTS) {
  for (sitename_i in all_sitenames) {
    # if the number of years spanned by the sitename is less than or equal to 2, skip it
    # ie if the difference between the max year and min year is less than or equal to 0 or 1, skip it
    
    # get the max and min years
    site_years <- ff_joined_with_boundaries %>%
      filter(sitename == sitename_i) %>%
      mutate(year = lubridate::year(close_date)) %>%
      summarize(min_year = min(year, na.rm = TRUE), max_year = max(year, na.rm = TRUE))
    
    if (site_years$max_year - site_years$min_year <= 1) {
      next
    }
    
    plot <- create_yearly_noncum_plot(sitename_i)
    filename <- paste0(sitename_i, "_yearly_noncum.png")
    
    # change slashes to dashes to avoid creating new folders
    filename <- str_replace_all(filename, "/", "-")
    filepath <- here("data_outputs", "ff_yearly_noncum_plots", filename)
    ggsave(filename = filepath, plot = plot, width = 8, height = 6)
  }
}
```




