
In this notebook, we'll load up non cumulative year-by-year acquisition data

We'll calculate the summary statistical moments for each
- mean
- standard deviation
- skew
- kurtosis


```{r}
# clear env
rm(list=ls())

# load libs
library(tidyverse)
library(here)
library(janitor)

# load dataset
# let's get the joined dataset in which the acquisitions have been associated
# with their bounding FF BOT projects
# data_raw/FFBOT_Spatial_Join_Acq-Proj_v3.csv

ff_joined_raw <- read_csv(here("data_raw", "FFBOT_Spatial_Join_Acq-Proj_v3.csv")) %>%
  clean_names()

```

Let's grab a subset we're interested in

```{r}
ff_joined_subset <- ff_joined_raw %>%
  select(close_date, shape_area, sitename) %>%
  rename(acquisition_acres = shape_area)
```

And let's also make a nice year field, since we only care about that. 
(We can actually just throw out the full date field)

```{r}
# Date format looks like
# 9/18/2001 0:00:00
ff_yearly_acquisitions <- ff_joined_subset %>%
  mutate(year = as.numeric(format(as.Date(close_date, format="%m/%d/%Y %H:%M:%S"), "%Y"))) %>% 
  select(-close_date)
```


## Obtaining probability functions

For each sitename, we want to get the year-by-year proportion of the total acquisition that was acquired in that year.
```{r}
ff_yearly_noncum_prop_acquired_by_site <- ff_yearly_acquisitions %>%
  group_by(sitename, year) %>%
  summarise(yearly_acquired = sum(acquisition_acres, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(sitename) %>%
  mutate(yearly_proportion_acquired = yearly_acquired / sum(yearly_acquired, na.rm = TRUE)) %>%
  ungroup()
```

## Num Distinct Years

Great now we have probability functions for each site.
Because we'll later want to do some filtering based on the number of distinct years thath acquisitions occurred within, let's also get a dataframe 
that indicates the number of distinct years that acquisitions occurred within for each site.

```{r}
n_distinct_years_by_site <- ff_yearly_acquisitions %>%
  group_by(sitename) %>%
  summarise(n_distinct_years = n_distinct(year)) %>%
  ungroup()
```


## Statistical Moments Helper Functions
Let's define some helper functions for obtaining statistical moments given a probability function

```{r}
# function to calculate mean given probability function
calculate_mean <- function(years, probs) {
  sum(years * probs)
}

# function to calculate standard deviation given probability function
calculate_sd <- function(years, probs) {
  mean <- calculate_mean(years, probs)
  sd <- sqrt(sum(probs * (years - mean)^2))
  return(sd)
}

# function to calculate skewness given probability function
calculate_skewness <- function(years, probs) {
  mean <- calculate_mean(years, probs)
  sd <- calculate_sd(years, probs)
  skewness <- sum(probs * ((years - mean) / sd)^3)
  return(skewness)
}

# function to calculate kurtosis given probability function
calculate_kurtosis <- function(years, probs) {
  mean <- calculate_mean(years, probs)
  sd <- calculate_sd(years, probs)
  kurtosis <- sum(probs * ((years - mean) / sd)^4)
  return(kurtosis)
}

```


## Computing Moments for each Site

Now we can compute the moments for each site using the helper functions we just defined.

```{r}
statistical_moments_df <- ff_yearly_noncum_prop_acquired_by_site %>%
  group_by(sitename) %>%
  summarise(
    mean_year = calculate_mean(year, yearly_proportion_acquired),
    sd_year = calculate_sd(year, yearly_proportion_acquired),
    skewness_year = calculate_skewness(year, yearly_proportion_acquired),
    kurtosis_year = calculate_kurtosis(year, yearly_proportion_acquired)
  ) %>%
  ungroup() %>%
  left_join(n_distinct_years_by_site, by = "sitename") # Join with the number of distinct years dataframe to get that info in the same dataframe
```


## Saving

Let's save our file in data_outputs with a meaningful name

```{r}
write_csv(statistical_moments_df, here("data_outputs", "FFBOT_Site_Statistical_Moments.csv"))
```
