
In this notebook, we'll load up the csv version for the non-cumulative plots created in 
ex03.1

We'll calculate the summary statistical moments for each
- mean
- standard deviation
- skew
- kurtosis

Later we'll do some clustering to see if we can find any patterns in the data. For now though, let's get these moments and get them
in a datafile we can use conveniently.

```{r}
# clear env
rm(list=ls())

# load libs
library(tidyverse)
library(here)
library(janitor)

# load dataset
# data_outputs/ff_yearly_noncum_prop_acquired_by_site.csv
ff_yearly_noncum_prop_acquired_by_site <- read_csv(here("data_outputs", "ff_yearly_noncum_proportion_acquired_by_site.csv")) %>%
  clean_names()

```

Let's define some helper functions for obtaining statistical moments given a probability function

```{r}
# function to calculate mean given probability function
calculate_mean <- function(years, probs) {
  sum(years * probs)
}

# function to calculate standard deviation given probability function
calculate_sd <- function(years, probs) {
  mean <- calculate_mean(years, probs)
  sd <- sqrt(sum(probs * (years - mean)^2))
  return(sd)
}

# function to calculate skewness given probability function
calculate_skewness <- function(years, probs) {
  mean <- calculate_mean(years, probs)
  sd <- calculate_sd(years, probs)
  skewness <- sum(probs * ((years - mean) / sd)^3)
  return(skewness)
}

# function to calculate kurtosis given probability function
calculate_kurtosis <- function(years, probs) {
  mean <- calculate_mean(years, probs)
  sd <- calculate_sd(years, probs)
  kurtosis <- sum(probs * ((years - mean) / sd)^4)
  return(kurtosis)
}

```

Now let's go through each of the sites and get the moments.

```{r}
statistical_moments_df <- data.frame(
  sitename = character(),
  mean = numeric(),
  sd = numeric(),
  skewness = numeric(),
  kurtosis = numeric(),
  stringsAsFactors = FALSE
)

all_sitenames <- unique(ff_yearly_noncum_prop_acquired_by_site$sitename)


for(site in all_sitenames) {
  # get vector of years and vector of probabilities for this site
  # Note the proportions do not add up to 1 -- since they refer to the 
  # proportion of the larger project bound yet to be completed. 
  # So in order to obtain a vector of probabilities, we need to get the total
  # proportion for the site and normalize by that.
  site_data <- ff_yearly_noncum_prop_acquired_by_site %>%
    filter(sitename == site) %>%
    arrange(year)
  
  total_proportion <- sum(site_data$yearly_proportion_acquired, na.rm = TRUE)
  
  site_data <- site_data %>% mutate(probability = yearly_proportion_acquired / total_proportion)
  
  years <- site_data$year
  probs <- site_data$probability
  
  # # to test, display years and probs vector to console
  # print(paste("Site:", site))
  # print("Years:")
  # print(years)
  # print("Probs:")
  # print(probs)
  # 
  
  # calculate moments
  mean <- calculate_mean(years, probs)
  sd <- calculate_sd(years, probs)
  skewness <- calculate_skewness(years, probs)
  kurtosis <- calculate_kurtosis(years, probs)
  
  # add to dataframe
  statistical_moments_df <- rbind(statistical_moments_df, data.frame(
    sitename = site,
    mean = mean,
    sd = sd,
    skewness = skewness,
    kurtosis = kurtosis,
    stringsAsFactors = FALSE
  ))
  
}




# compute excess_kurtosis (just kurtosis - 3, since normal distribution has kurtosis of 3)
statistical_moments_df <- statistical_moments_df %>%
  mutate(excess_kurtosis = kurtosis - 3)

```


Let's save as a datafile in data_outputs dir

```{r}
write_csv(statistical_moments_df, here("data_outputs", "ff_yearly_noncum_statistical_moments_by_site.csv"))
```
