
```{r}
# clear env
rm(list=ls())

# load libs
library(tidyverse)
library(here)
library(janitor)

# load dataset
# data_outputs/ff_yearly_noncum_statistical_moments_by_site.csv
ff_yearly_noncum_statistical_moments_by_site <- read_csv(here("data_outputs", "ff_yearly_noncum_statistical_moments_by_site.csv")) %>%
  clean_names()
```



First we need to normalize our statistical moments data before using them for clustering.

Let's visualize the distributions of each first.

## Visualizing statistical moments

**Mean**

```{r}
ggplot(ff_yearly_noncum_statistical_moments_by_site, aes(x = mean)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of Mean", x = "Mean", y = "Count")
```
**SD**

```{r}
ggplot(ff_yearly_noncum_statistical_moments_by_site, aes(x = sd)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of SD", x = "SD", y = "Count")
```

**Skewness**

```{r}
ggplot(ff_yearly_noncum_statistical_moments_by_site, aes(x = skewness)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of Skewness", x = "Skewness", y = "Count")

```
**Kurtosis**

```{r}
ggplot(ff_yearly_noncum_statistical_moments_by_site, aes(x = kurtosis)) +
  geom_histogram(bins = 30) +
  theme_minimal() +
  labs(title = "Distribution of Kurtosis", x = "Kurtosis", y = "Count")


```

Ok we've got some crazy crazy outliers for kurtosis and skew
Both for the same project:
Etoniah/Cross Florida Greenway Florida Forever BOT Project

These extreme values could warrant some extra investigation.



For now, let's get the clustering code in here so we have it.


## Normalizing the data

Let's obtain z scores

```{r}
moments_z <- ff_yearly_noncum_statistical_moments_by_site %>%
  select(mean, sd, skewness, excess_kurtosis) %>%
  scale() %>%
  as.data.frame() %>%
  set_names(c("mean_z", "sd_z", "skewness_z", "exc_kurtosis_z"))

```



## Clustering

Let's do k-means

```{r}
set.seed(123) # for reproducibility

k <- 4 # number of clusters

kmeans_result <- kmeans(moments_z, centers = k)

# Add cluster assignments back to original data
sites_clustered <- ff_yearly_noncum_statistical_moments_by_site %>%
  mutate(cluster = kmeans_result$cluster)

```


## Analyzing clusters

Let's see how the clusters differ in terms of their mean, sd, skewness, and kurtosis

```{r}
cluster_summary <- sites_clustered %>%
  group_by(cluster) %>%
  summarize(
    mean_mean = mean(mean),
    mean_sd = mean(sd),
    mean_skewness = mean(skewness),
    mean_exc_kurtosis = mean(excess_kurtosis)
  )

print(cluster_summary)
```

How can we get a more complete story besides just getting the means of each cluster? The answer of course is to visualize the clusters in a 2D space using PCA or t-SNE. Let's do PCA for now.

```{r}
# Perform PCA
pca_result <- prcomp(moments_z, center = TRUE, scale. = TRUE)

# Get PCA scores
pca_scores <- as.data.frame(pca_result$x) %>%
  set_names(c("PC1", "PC2", "PC3", "PC4"))

# Add cluster assignments to PCA scores
pca_scores <- pca_scores %>%
  mutate(cluster = kmeans_result$cluster)

# Visualize clusters in PCA space
ggplot(pca_scores, aes(x = PC1, y = PC2, color = factor(cluster))) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "K-means Clusters in PCA Space", x = "Principal Component 1", y = "Principal Component 2", color = "Cluster")

```

```{r}
# PCA graph with vectors
pca_loadings <- as.data.frame(pca_result$rotation) %>%
  set_names(c("PC1", "PC2", "PC3", "PC4")) %>%
  rownames_to_column(var = "variable")


ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  geom_point(size = 3) +
  geom_segment(data = pca_loadings, aes(x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5), arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text(data = pca_loadings, aes(x = PC1 * 5, y = PC2 * 5, label = variable), vjust = -0.5) +
  theme_minimal() +
  labs(title = "K-means Clusters in PCA Space with Loadings", x = "Principal Component 1", y = "Principal Component 2", color = "Cluster")
```



```{r}
# Scree plot
pca_var <- pca_result$sdev^2

pca_var_perc <- pca_var / sum(pca_var) * 100
pca_var_df <- data.frame(PC = paste0("PC", 1:length(pca_var_perc)), Variance = pca_var_perc)
ggplot(pca_var_df, aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Scree Plot", x = "Principal Component", y = "Percentage of Variance Explained")
```





