

I randomly chose 3 examples from each of 4 clusters from previous notebook


**Cluster 1** (SOUTH)
- Sand Mountain Florida Forever BOT Project
- Lake Wales Ridge Ecosystem Florida Forever BOT Project - Carter Creek
- Brevard Coastal Scrub Ecosystem Florida Forever BOT Project - Scrub Jay Refugia - Valkaria
- Florida Keys Ecosystem Florida Forever BOT Project - Islamorada/Plantation Key

**Cluster 2** NORTH
- Brevard Coastal Scrub Ecosystem Florida Forever BOT Project - Grissom Parkway
- Corkscrew Regional Ecosystem Watershed Florida Forever BOT Project
- Osceola Pine Savannas Florida Forever BOT Project
- Wekiva-Ocala Greenway Florida Forever BOT Project - Wekiva-Ocala Connector West


**Cluster 3** EAST
- Garcon Ecosystem Florida Forever BOT Project
- Rainbow River Corridor Florida Forever BOT Project
- Red Hills Conservation Florida Forever BOT Project
- Pumpkin Hill Creek Florida Forever BOT Project


**Cluster 4** WEST
- Middle Chipola River Florida Forever BOT Project
- Brevard Coastal Scrub Ecosystem Florida Forever BOT Project - Micco Expansion
- Belle Meade Florida Forever BOT Project
- Florida Keys Ecosystem Florida Forever BOT Project - Hammocks of the Lower Keys - Big Torch Key


Now I'm going to plot the year by year non-cumulative acquired acrage of each of these sites

```{r}
# clear env
rm(list=ls())

# load libs
library(tidyverse)
library(here)

# load dataset
ff_joined <- read_csv(here("data_raw", "FFBOT_Spatial_Join_Acq-Proj.csv")) %>% 
  clean_names()

```

```{r}
# grab the subset of the data for the 12 sites of interest
sites_of_interest <- c("Sand Mountain Florida Forever BOT Project",
                       "Lake Wales Ridge Ecosystem Florida Forever BOT Project - Carter Creek",
                       "Brevard Coastal Scrub Ecosystem Florida Forever BOT Project - Scrub Jay Refugia - Valkaria",
                       "Florida Keys Ecosystem Florida Forever BOT Project - Islamorada/Plantation Key",
                       "Brevard Coastal Scrub Ecosystem Florida Forever BOT Project - Grissom Parkway",
                       "Corkscrew Regional Ecosystem Watershed Florida Forever BOT Project",
                       "Osceola Pine Savannas Florida Forever BOT Project",
                       "Wekiva-Ocala Greenway Florida Forever BOT Project - Wekiva-Ocala Connector West",
                       "Garcon Ecosystem Florida Forever BOT Project",
                       "Rainbow River Corridor Florida Forever BOT Project",
                       "Red Hills Conservation Florida Forever BOT Project",
                       "Pumpkin Hill Creek Florida Forever BOT Project",
                       "Middle Chipola River Florida Forever BOT Project",
                       "Brevard Coastal Scrub Ecosystem Florida Forever BOT Project - Micco Expansion",
                       "Belle Meade Florida Forever BOT Project",
                       "Florida Keys Ecosystem Florida Forever BOT Project - Hammocks of the Lower Keys - Big Torch Key")

ff_joined_subset <- ff_joined %>%
  filter(sitename %in% sites_of_interest) %>%
  select(close_date, shape_area, sitename) %>%
  rename(acquisition_acres = shape_area) %>%
  mutate(year = as.numeric(format(as.Date(close_date, format="%m/%d/%Y %H:%M:%S"), "%Y"))) %>% 
  select(-close_date)
```

plot year by year non-cumulative acquired acreage for each site seperately


```{r}
# Define helper function to make a plot for one specific site passed in as a parameter
# plot_site_acquisition <- function(site_name) {
#   site_data <- ff_joined_subset %>%
#     filter(sitename == site_name) %>%
#     group_by(year) %>%
#     summarise(yearly_acquired = sum(acquisition_acres, na.rm = TRUE)) %>%
#     ungroup()
#   
#   ggplot(site_data, aes(x = year, y = yearly_acquired)) +
#     geom_bar(stat = "identity", fill = "steelblue") +
#     labs(title = site_name, x = "Year", y = "Acres Acquired") +
#     theme_minimal()
# }

# Redifene, but ensure that there is exactly one tick on the x axis for each year. One bin per year
plot_site_acquisition <- function(site_name) {
  site_data <- ff_joined_subset %>%
    filter(sitename == site_name) %>%
    group_by(year) %>%
    summarise(yearly_acquired = sum(acquisition_acres, na.rm = TRUE)) %>%
    ungroup()
  
  # ggplot(site_data, aes(x = year, y = yearly_acquired)) +
  #   geom_bar(stat = "identity", fill = "steelblue") +
  #   scale_x_continuous(breaks = seq(min(site_data$year), max(site_data$year), by = 1)) +
  #   labs(title = site_name, x = "Year", y = "Acres Acquired") +
  #   theme_minimal()
  # 
  # Rotate Years by 45 degrees
  ggplot(site_data, aes(x = year, y = yearly_acquired)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    scale_x_continuous(breaks = seq(min(site_data$year), max(site_data$year), by = 1)) +
    labs(title = site_name, x = "Year", y = "Acres Acquired") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r}
# Now we can loop through the sites of interest and make a plot for each one
for (site in sites_of_interest) {
  print(plot_site_acquisition(site))
}
```





