
```{r}
# clear env
rm(list=ls())

# load libs
library(tidyverse)
library(here)
library(janitor)

# load dataset
# data_outputs/FFBOT_Site_Statistical_Moments.csv
noncum_moments <- read_csv(here("data_outputs", "FFBOT_Site_Statistical_Moments.csv")) %>%
  clean_names()
```

Our columns are
[1] "sitename"         "mean_year"        "sd_year"          "skewness_year"    "kurtosis_year"    "n_distinct_years"


## Filtering (sites with sufficient data)

We aren't interested in all data, let's filter to get those whose n_distinct_years is greater than 2

```{r}
noncum_moments_filtered <- noncum_moments %>%
  filter(n_distinct_years > 2)
```

### The CRAZY outiler

Dickerson Bay/Bald Point Florida Forever BOT Project

```{r}
noncum_moments_filtered <- noncum_moments_filtered %>%
  filter(sitename != "Dickerson Bay/Bald Point Florida Forever BOT Project")
```




## Normalizing

We need to normalize our statistical moments data before using them for clustering. Let's visualize the distributions of each first. Let's do it by taking z_scores for each.

```{r}
moments_z <- noncum_moments_filtered %>%
  select(mean_year, sd_year, skewness_year, kurtosis_year) %>%
  scale() %>%
  as.data.frame() %>%
  set_names(c("mean_z", "sd_z", "skewness_z", "kurtosis_z"))
```


## PCA

Let's do some PCA to see if we can see this in 2D and guess a good cluster number.

```{r}
pca_result <- prcomp(moments_z, center = TRUE, scale. = TRUE)
pca_data <- as.data.frame(pca_result$x) %>%
  select(PC1, PC2)

ggplot(pca_data, aes(x = PC1, y = PC2)) +
  geom_point() +
  theme_minimal() +
  labs(title = "PCA of Statistical Moments", x = "Principal Component 1", y = "Principal Component 2")

```

## Clustering (k-means)

```{r}
set.seed(123) # for reproducibility

# Determine optimal number of clusters using the elbow method
wss <- sapply(1:10, function(k) {
  kmeans(moments_z, centers = k, nstart = 25)$tot.withinss
})
```

Let's see the elbow plot

```{r}
elbow_data <- data.frame(k = 1:10, wss = wss)
ggplot(elbow_data, aes(x = k, y = wss)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Elbow Method for Determining Optimal Clusters", x = "Number of Clusters (k)", y = "Total Within-Cluster Sum of Squares")
```
There is no clearly distinct "elbow".
There is maybe some signal at k=7?

Let's use that for now.

```{r}
k <- 7 # number of clusters

kmeans_result <- kmeans(moments_z, centers = k, nstart = 25)
# Add cluster assignments back to original data
sites_clustered <- noncum_moments_filtered %>%
  mutate(cluster = kmeans_result$cluster)

```

## Making Sense of the Clusters

```{r}
ggplot(sites_clustered, aes(x = mean_year, y = sd_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on Mean and SD", x = "Mean", y = "SD", color = "Cluster")
```

```{r}

ggplot(sites_clustered, aes(x = skewness_year, y = kurtosis_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on Skewness and Kurtosis (Log Scale)", x = "Skewness", y = "Kurtosis", color = "Cluster") +
  scale_x_log10() +
  scale_y_log10()
```

### LDA

Let's use LDA to comprehensively understand the clusters and which variables are most important for distinguishing them.

```{r}
# Do not load MASS
# instead reference it with the :: syntax to avoid conflicts with dplyr

lda_result <- MASS::lda(cluster ~ mean_year + sd_year + skewness_year + kurtosis_year, data = sites_clustered)
lda_data <- as.data.frame(predict(lda_result)$x) %>%
  bind_cols(sites_clustered %>% select(cluster))
ggplot(lda_data, aes(x = LD1, y = LD2, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "LDA of Clusters", x = "Linear Discriminant 1", y = "Linear Discriminant 2", color = "Cluster")

```

Now let's make sense of the discriminants

```{r}
lda_coefficients <- as.data.frame(lda_result$scaling) %>%
  rownames_to_column(var = "variable")

# display
lda_coefficients
```

Create LDA plot again but with vectors for the coefficients

```{r}
# scaled vectors by 10 for visibility
ggplot(lda_data, aes(x = LD1, y = LD2, color = factor(cluster))) +
  geom_point() +
  geom_segment(data = lda_coefficients, aes(x = 0, y = 0, xend = LD1 * 10, yend = LD2 * 10), arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text(data = lda_coefficients, aes(x = LD1 * 3.5, y = LD2 * 3.5, label = variable), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "LDA of Clusters with Scaled Coefficients", x = "Linear Discriminant 1", y = "Linear Discriminant 2", color = "Cluster")
```

Wow, LDA 2 by itself defines a very significant amount of distinction between the groups

Skewness, Mean, and SD are all closely related. Kurtosis, meanwhile, seems more independent.

Surprisingly, lower skewness is correlated with lower mean year, the opposite of what one would expect for a normal distribution.







