
```{r}
# clear env
rm(list=ls())

# load libs
library(tidyverse)
library(here)
library(janitor)


SAVE_CLUSTER_RESULTS <- FALSE

```


```{r}
# load dataset
# data_outputs/FFBOT_Site_Statistical_Moments.csv
noncum_moments <- read_csv(here("data_outputs", "FFBOT_Site_Statistical_Moments.csv")) %>%
  clean_names()
```

Our columns are
[1] "sitename"         "mean_year"        "sd_year"          "skewness_year"    "kurtosis_year"    "n_distinct_years"


## Filtering (sites with sufficient data)

We aren't interested in all data, let's filter to get those whose n_distinct_years is greater than 2

```{r}
noncum_moments_filtered <- noncum_moments %>%
  filter(n_distinct_years > 2)
```

### Identifying outliers

```{r}
# noncum_moments_filtered <- noncum_moments_filtered %>%
#   filter(sitename != "St. Joe Timberland Florida Forever BOT Project - Wakulla Springs Protection Zone")
```


Let's get a list of outliers for each of the 4 moments we've calculated
mean_year, sd_year, skewness_year, kurtosis_year

**Mean Year Outliers**
```{r}
# get outliers for mean_year
noncum_moments_filtered %>%
  filter(mean_year > mean(mean_year) + 2 * sd(mean_year) | mean_year < mean(mean_year) - 2 * sd(mean_year)) %>%
  select(sitename, mean_year)
```

**SD Year OUtliers**
```{r}
noncum_moments_filtered %>%
  filter(sd_year > mean(sd_year) + 2 * sd(sd_year) | sd_year < mean(sd_year) - 2 * sd(sd_year)) %>%
  select(sitename, sd_year)
```

**Skewness Outilers**
```{r}
noncum_moments_filtered %>%
  filter(skewness_year > mean(skewness_year) + 2 * sd(skewness_year) | skewness_year < mean(skewness_year) - 2 * sd(skewness_year)) %>%
  select(sitename, skewness_year)

```

**Kurtossis Outliers**
```{r}
noncum_moments_filtered %>%
  filter(kurtosis_year > mean(kurtosis_year) + 2 * sd(kurtosis_year) | kurtosis_year < mean(kurtosis_year) - 2 * sd(kurtosis_year)) %>%
  select(sitename, kurtosis_year)
```



### Dealing with outliers



Identified Outliers:


St. Joe Timberland Florida Forever BOT Project - Wakulla Springs Protection Zone
Florida's First Magnitude Springs Florida Forever BOT Project - Manatee Springs
Northeast Florida Blueway Florida Forever BOT Project


Filter them out
```{r}
noncum_moments_filtered <- noncum_moments_filtered %>%
  filter(!sitename %in% c("St. Joe Timberland Florida Forever BOT Project - Wakulla Springs Protection Zone",
                          "Florida's First Magnitude Springs Florida Forever BOT Project - Manatee Springs",
                          "Northeast Florida Blueway Florida Forever BOT Project"))

```




## Normalizing

We need to normalize our statistical moments data before using them for clustering. Let's visualize the distributions of each first. Let's do it by taking z_scores for each.

```{r}
moments_z <- noncum_moments_filtered %>%
  select(mean_year, sd_year, skewness_year, kurtosis_year) %>%
  scale() %>%
  as.data.frame() %>%
  set_names(c("mean_z", "sd_z", "skewness_z", "kurtosis_z"))
```


## PCA

Let's do some PCA to see if we can see this in 2D and guess a good cluster number.

```{r}
pca_result <- prcomp(moments_z, center = TRUE, scale. = TRUE)
pca_data <- as.data.frame(pca_result$x) %>%
  select(PC1, PC2)

pca_data_with_vectors <- as.data.frame(pca_result$x) %>%
  select(PC1, PC2)

ggplot(pca_data_with_vectors, aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_segment(data = as.data.frame(pca_result$rotation) %>%
                 rownames_to_column("variable") %>%
                 select(variable, PC1, PC2),
               aes(x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5), # scale vectors for better visualization
               arrow = arrow(length = unit(0.2, "cm")), color = "red") +
  geom_text(data = as.data.frame(pca_result$rotation) %>%
              rownames_to_column("variable") %>%
              select(variable, PC1, PC2),
            aes(x = PC1 * 5.5, y = PC2 * 5.5, label = variable), color = "red", size = 3) +
  theme_minimal() +
  labs(title = "PCA of Statistical Moments with Vectors and Labels", x = "Principal Component 1", y = "Principal Component 2")
```

Let's get a scree plot

```{r}
scree_data <- data.frame(PC = paste0("PC", 1:length(pca_result$sdev)), Variance = pca_result$sdev^2 / sum(pca_result$sdev^2))
ggplot(scree_data, aes(x = PC, y = Variance)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Scree Plot", x = "Principal Component", y = "Proportion of Variance Explained")
```



## Clustering (k-means)

```{r}
set.seed(123) # for reproducibility

# Determine optimal number of clusters using the elbow method
wss <- sapply(1:10, function(k) {
  kmeans(moments_z, centers = k, nstart = 25)$tot.withinss
})
```

Let's see the elbow plot

```{r}
elbow_data <- data.frame(k = 1:10, wss = wss)
ggplot(elbow_data, aes(x = k, y = wss)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Elbow Method for Determining Optimal Clusters", x = "Number of Clusters (k)", y = "Total Within-Cluster Sum of Squares")
```
There is no clearly distinct "elbow".
There is maybe some signal at k=7?

Let's use that for now.

```{r}
k <- 4 # number of clusters

kmeans_result <- kmeans(moments_z, centers = k, nstart = 25)
# Add cluster assignments back to original data
sites_clustered <- noncum_moments_filtered %>%
  mutate(cluster = kmeans_result$cluster)

```

## Making Sense of the Clusters

```{r}
ggplot(sites_clustered, aes(x = mean_year, y = sd_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on Mean and SD", x = "Mean", y = "SD", color = "Cluster")
```


```{r}

ggplot(sites_clustered, aes(x = skewness_year, y = kurtosis_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on Skewness and Kurtosis", x = "Skewness", y = "Kurtosis", color = "Cluster")

# ggplot(sites_clustered, aes(x = skewness_year, y = kurtosis_year, color = factor(cluster))) +
#   geom_point() +
#   theme_minimal() +
#   labs(title = "Clusters of Sites Based on Skewness and Kurtosis (Log Scale)", x = "Skewness", y = "Kurtosis", color = "Cluster") +
#   scale_x_log10() +
#   scale_y_log10()


```

Let's also get plots that compare other combos

Like mean vs skewness, mean vs kurtosis, sd vs skewness, sd vs kurtosis

```{r}
ggplot(sites_clustered, aes(x = mean_year, y = skewness_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on Mean and Skewness", x = "Mean", y = "Skewness", color = "Cluster")
```

```{r}
ggplot(sites_clustered, aes(x = mean_year, y = kurtosis_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on Mean and Kurtosis", x = "Mean", y = "Kurtosis", color = "Cluster")
```

```{r}
ggplot(sites_clustered, aes(x = sd_year, y = skewness_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on SD and Skewness", x = "SD", y = "Skewness", color = "Cluster")
```

```{r}
ggplot(sites_clustered, aes(x = sd_year, y = kurtosis_year, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "Clusters of Sites Based on SD and Kurtosis", x = "SD", y = "Kurtosis", color = "Cluster")
```


### PC Plot with color-coded clusters

Let's see that Principal Component Plot again, but now with color-coded clusters.
We'll keep the vectors for PCs and all.
```{r}
pca_data_with_clusters <- pca_data %>%
  bind_cols(sites_clustered %>% select(cluster))
ggplot(pca_data_with_clusters, aes(x = PC1, y = PC2, color = factor(cluster))) +
  geom_point() +
  geom_segment(data = as.data.frame(pca_result$rotation) %>%
                 rownames_to_column("variable") %>%
                 select(variable, PC1, PC2),
               aes(x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5), # scale vectors for better visualization
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text(data = as.data.frame(pca_result$rotation) %>%
              rownames_to_column("variable") %>%
              select(variable, PC1, PC2),
            aes(x = PC1 * 5.5, y = PC2 * 5.5, label = variable), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "PCA of Statistical Moments with Clusters and Vectors", x = "Principal Component 1", y = "Principal Component 2", color = "Cluster")

# ggplot(pca_data_with_clusters, aes(x = PC1, y = PC2, color = factor(cluster))) +
#   geom_point() +
#   theme_minimal() +
#   labs(title = "PCA of Statistical Moments with Clusters", x = "Principal Component 1", y = "Principal Component 2", color = "Cluster")
```

## Seeing site labels

Let's see this exact plot again, but now labeling each dot with the corresponding site
```{r}
# first, sitename needs to be added to pca_data_with_clusters
# we can do this by taking the sitename col from noncum_moments_filtered and adding it to pca_data_with_clusters
pca_data_with_clusters_with_sites <- pca_data_with_clusters %>%
  bind_cols(noncum_moments_filtered %>% select(sitename))

ggplot(pca_data_with_clusters_with_sites, aes(x = PC1, y = PC2, color = factor(cluster))) +
  geom_point() +
  geom_text(aes(label = sitename), size = 3, vjust = -1) +
  geom_segment(data = as.data.frame(pca_result$rotation) %>%
                 rownames_to_column("variable") %>%
                 select(variable, PC1, PC2),
               aes(x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5), # scale vectors for better visualization
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text(data = as.data.frame(pca_result$rotation) %>%
              rownames_to_column("variable") %>%
              select(variable, PC1, PC2),
            aes(x = PC1 * 5.5, y = PC2 * 5.5, label = variable), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "PCA of Statistical Moments with Clusters and Vectors (Labeled)", x = "Principal Component 1", y = "Principal Component 2", color = "Cluster")
```


Let's see that plot again, but only show 10% of the labels
(Can keep regenerating this to try to see the labels of specific dots in the cluster)

```{r}

pca_data_with_clusters_with_sites_sampled <- pca_data_with_clusters_with_sites %>%
  sample_frac(0.05) # keep only 10% of the rows (or some other proportion)
ggplot(pca_data_with_clusters_with_sites_sampled, aes(x = PC1, y = PC2, color = factor(cluster))) +
  geom_point() +
  geom_text(aes(label = sitename), size = 3, vjust = -1) +
  geom_segment(data = as.data.frame(pca_result$rotation) %>%
                 rownames_to_column("variable") %>%
                 select(variable, PC1, PC2),
               aes(x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5), # scale vectors for better visualization
               arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text(data = as.data.frame(pca_result$rotation) %>%
              rownames_to_column("variable") %>%
              select(variable, PC1, PC2),
            aes(x = PC1 * 5.5, y = PC2 * 5.5, label = variable), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "PCA of Statistical Moments with Clusters and Vectors (Labeled Sample)", x = "Principal Component 1", y = "Principal Component 2", color = "Cluster")
```



### LDA

Let's use LDA to comprehensively understand the clusters and which variables are most important for distinguishing them.

```{r}
# Do not load MASS
# instead reference it with the :: syntax to avoid conflicts with dplyr

lda_result <- MASS::lda(cluster ~ mean_year + sd_year + skewness_year + kurtosis_year, data = sites_clustered)
lda_data <- as.data.frame(predict(lda_result)$x) %>%
  bind_cols(sites_clustered %>% select(cluster))
ggplot(lda_data, aes(x = LD1, y = LD2, color = factor(cluster))) +
  geom_point() +
  theme_minimal() +
  labs(title = "LDA of Clusters", x = "Linear Discriminant 1", y = "Linear Discriminant 2", color = "Cluster")

```

Now let's make sense of the discriminants

```{r}
lda_coefficients <- as.data.frame(lda_result$scaling) %>%
  rownames_to_column(var = "variable")

# display
lda_coefficients
```

Create LDA plot again but with vectors for the coefficients

```{r}
# scaled vectors by 10 for visibility
ggplot(lda_data, aes(x = LD1, y = LD2, color = factor(cluster))) +
  geom_point() +
  geom_segment(data = lda_coefficients, aes(x = 0, y = 0, xend = LD1 * 10, yend = LD2 * 10), arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text(data = lda_coefficients, aes(x = LD1 * 3.5, y = LD2 * 3.5, label = variable), color = "black", size = 3) +
  theme_minimal() +
  labs(title = "LDA of Clusters with Scaled Coefficients", x = "Linear Discriminant 1", y = "Linear Discriminant 2", color = "Cluster")
```

Wow, LDA 2 by itself defines a very significant amount of distinction between the groups

Skewness, Mean, and SD are all closely related. Kurtosis, meanwhile, seems more independent.

Surprisingly, lower skewness is correlated with lower mean year, the opposite of what one would expect for a normal distribution.

## Save Clustering Results
Let's save our cluster results as a dataframe so that we can use them in other scripts.

We already have a dataset that has all the statistical momentns (that's what we had loaded in at the start of the script)

So it would make sense for us to save a dataset that only consists of the cluster assignments paired with the primary keys for the sites (the sitenames)

```{r}
# write to data_outputs/ffbot_sites_clustered

if(SAVE_CLUSTER_RESULTS) {

  save_path <- here::here("data_outputs", "ffbot_sites_clustered.csv")
  
  sites_clustered %>% select(sitename, cluster) %>%
    write_csv(save_path)

}
  

```




